Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-22 doc/asplos13/grappa.tex

What makes Grappa's tasks special is that when they perform
long-latency operations they can give up control of their core,
allowing the core to remain busy while waiting for the operation to
complete. This most often done implicitly inside a call into Grappa's
API, but it can also be done explicitly by the programmer using the
calls shown in Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase
    reads and writes, but include that we provide a way to overlap
    computation}

All Grappa scheduling is done in userspace, and minimal state is
saved, in order to minimize the context switch time. In our
experiments, we saw context switch times as low as \TODO{40ns}.
This lightweight context switching is Grappa's key enabling
feature. It gives us the ability to tolerate latency. Given that
ability, we are able to trade latency for throughput: by {\em
  increasing} latency in key components in the system we are able to
increase our aggregate random access bandwidth, our synchronization
bandwidth, and our ability to tolerate load.

@@

Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. The first is explicit task
spawns. When the programmer identifies work that can be done in
parallel, the work may be wrapped up in a function and queued with its
arguments for later execution using a \texttt{spawn}. Sometimes the
programmer may want to spawn a task on a specific core in the system,
or at the home core of a particular memory location. Grappa provides a
\texttt{spawn\_on} call for this purpose; the task is immediately
enqueued on the specified core and is not available to other cores.

The next method for expressing parallelism is a parallel for loop, where the
number of iterations must be known at loop entry. The programmer specifies a
function pointer along with start and end indices and an optional threshold to
control parallel overhead. Grappa does {\em recursive decomposition} of
iterations, as in the implementation of Cilk's cilk\_for construct \TODO{cite
cilk\_for implementation: could only find slides page6 of
http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-Cilk++.pdf}.
It generates a logarithmically-deep tree of tasks, stopping to execute the
loop body when the number of iterations is below the threshold (``grainsize''
in Cilk). \TODO{we may want to make our classification purer: the for loop is
not Grappa's core, and is built atop tasks as a useful construct; in some
sense the only way to express parallelism in Grappa is tasks and perhaps
call\_on. -BM}

Finally, a programmer may want to run a small piece of code on a
particular core in the system without waiting for execution resources
to be available. Grappa provides the \texttt{call\_on} call for this
purpose.

@@

\paragraph{Explicit caching}
With Grappa's explicit caching support, applications can instruct Grappa to fetch a global
pointer of any length and return a local pointer to a cached copy of
the global memory. Under the hood, Grappa performs the mechanics of
gathering chunks of data from multiple system nodes and presenting a
conventional appearing linear block of memory as a pointer into a
cache. Grappa cache operations have the usual read-only and read-write
variants, along with a write-only variant used to initialize data
structures. Cache operations exploit spatial locality by reducing the number of small network
messages accessing contiguous data. Languages for distributed
shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
coalesces struct and array accesses into 
remote get/put \cite{Chen:2005}, or Fortran D compiler's message
vectorization that hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a
mechanism for exploiting temporal locality by operating on the data locally.

\paragraph{Delegate operations}
When the access pattern has low-locality, it is more efficient
to modify the data on its home core rather than bringing a copy to the
requesting core and returning it after modification. Delegate
operations provide this capability. Applications can dispatch
computation to be performed on individual machine-word sized chunks of
global memory to the memory system itself (e.g.,
\emph{fetch-and-add}). \TODO{should we bring up delegation in the intro?}

Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the primary synchronization method in
Grappa. Each piece of global memory is paired with a single core in
the system; all accesses to that memory are done by that core. This
allows delegate operations to be performed atomically on the core
simply by making sure no other code interleaves; since context
switches in Grappa are cooperative, this is trivial. Thus, delegate
operations are able to provide atomic semantics to memory owned by one core without using atomic
operations. Using delegation to implement isolation for any subset of
memory has been explored in \cite{delegated:oopsla11}.

>>>>

Grappa's tasks give up control of their core when when they perform
long-latency operations, allowing the processor to remain busy while
waiting for the operation to complete. This is most often done
implicitly inside a call into Grappa's API, but it can also be done
explicitly by the programmer using the calls shown in
Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase reads and
writes, but include that we provide a way to overlap computation}

All Grappa scheduling is done in userspace, and minimal state is saved,
in order to minimize the context switch time. In our experiments, we saw
context switch times as low as \checkme{40ns}. This lightweight context
switching is Grappa's key enabling feature. It gives us the ability to
tolerate latency. Given that ability, we are able to \textbf{trade
latency for throughput}: by {\em increasing} latency in key components
in the system we are able to increase our aggregate random access
bandwidth, our synchronization bandwidth, and our ability to tolerate
load.

@@

Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. The first is explicit task
spawns. When the programmer identifies work that can be done in
parallel, the work may be wrapped up in a function and queued with its
arguments for later execution using a \texttt{spawn}. If the
programmer wishes to spawn a task on a specific core in the system,
or at the home core of a particular memory location, Grappa provides a
\texttt{spawn\_on} call for this purpose.

The next method for expressing parallelism is a parallel for loop, where
the number of iterations must be known at loop entry. The programmer
specifies a function pointer along with start and end indices and an
optional threshold to control parallel overhead. Grappa does {\em
recursive decomposition} of iterations, similar to Cilk's cilk\_for
construct~\cite {cilkforimplementation} \comment{could only find slides
page6 of
http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-
Cilk++.pdf    ---- Good enough.  Cite it.  -Mark}.  It generates a
logarithmically-deep tree of tasks, stopping to execute the loop body
when the number of iterations is below the required threshold. Finally,
a programmer may want to run a small piece of code on a particular core
in the system without waiting for execution resources to be available.
Grappa provides the \texttt{call\_on} call for this purpose.

@@

\paragraph{Explicit caching} With Grappa's explicit caching support,
applications can instruct Grappa to fetch a global pointer of any length
and return a local pointer to a cached copy of the global memory. Under
the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear
block of memory as a pointer into a cache. Grappa cache operations have
the usual read-only and read-write variants, along with a write-only
variant used to initialize data structures. Cache operations exploit
spatial locality by reducing the number of small network messages
accessing contiguous data. Languages for distributed shared memory
systems have done optimizations to achieve a similar goal. For example,
the UPC compiler coalesces struct and array accesses into remote get/put
\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
additionally provides a mechanism for exploiting temporal locality by
operating on the data locally.

\paragraph{Delegate operations}
When the access pattern has low-locality, it is more efficient
to modify the data on its home core rather than bringing a copy to the
requesting core and returning it after modification. Delegate
operations provide this capability. Applications can dispatch
computation to be performed on individual machine-word sized chunks of
global memory to the memory system itself (e.g.,
\emph{fetch-and-add}). \comment{should we bring up delegation in the intro?}

Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the
primary synchronization method in Grappa. Each piece of global memory is
paired with a single core in the system; all accesses to that actual
memory location are done by that core. This allows delegate operations
to be performed atomically on the core simply by making sure no other
code interleaves their operation; and, since context switches in Grappa
are cooperative, this is trivial. Thus, delegate operations are able to
provide atomic semantics to memory owned by one core without using
atomic operations. Using delegation to implement isolation for any
subset of memory has been explored in \cite{delegated:oopsla11}.

### 2012-07-22 doc/asplos13/intro.tex

What makes our goal challenging is that irregular applications do not readily exploit the features of commodity systems.  This is true even for single node systems.  It is not atypical for any given task's data references to be spread randomly across the entire memory of the system. Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.  Furthermore, the addresses of references are often determined in the task only just before their values are used.  Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.  Because commodity processors' pipeline utilization relies on spatial and temporal locality -- or, at least the ability to predict in advance what is needed -- they stall often when executing irregular applications.

On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.  Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.  Injection rate into the networks is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.  While some irregular applications can be restructured to better exploit locality, aggregate requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.

An idealized alternative to demanding this expertise of programmers was provided by the fully custom Tera MTA-2~\cite{mta-2} system.  The MTA was well-suited to extremely irregular applications: it had a large distributed shared memory with no caches, relying on multithreaded processors to tolerate memory latency.  On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts.  Any application with sufficient thread parallelism would fully utilize the processors independent of the memory access pattern (modulo hot-spots).  The network was designed with an single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle.  Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single threaded performance, making it a commercial failure.  The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.

In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.  This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating with inexpensive mass market hardware components the approach taken by Tera. We exploit the increased aggregate instruction rate per socket relative to chip bandwidth.  That is, where locality friendly computations make use of instructions to get work done in cache, we use what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.  Ultimately, the opportunity is to cover the spectrum of irregular to regular computation:  where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.  Thus we make best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.

>>>>

What makes our goal challenging is that irregular applications do not readily exploit the features of commodity systems.  This is true even for single node systems.  It is not atypical for any given task's data references to be spread randomly across the entire memory of the system. Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.  Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications.  As a consequence, commodity microprocessors stall often when executing irregular applications.

On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.  Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.  Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.  While some irregular applications can be restructured to better exploit locality, aggregating requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.

An alternative to demanding programmers manage locality in their applications, was provided by the fully custom Tera MTA-2~\cite{mta-2} system.  The MTA was well-suited to extremely irregular applications: it had a large distributed shared memory with no caches, relying on multithreaded processors to tolerate memory latency.  On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts.  Any application with sufficient thread parallelism would fully utilize the processors independent of the memory access pattern (modulo hot-spots).  The network was designed with an single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle.  Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single threaded performance, making it a commercial failure.  The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.

In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.  This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating with inexpensive mass market hardware components the approach taken by Tera. We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.  Ultimately, the opportunity is to cover the spectrum of irregular to regular computation:  where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.  Thus we make best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.
